Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany

### 2025-05-28 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
add qmd file

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[runningheads]{llncs}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{2}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[activate=false]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\makeatletter

\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter

\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany},
  pdfauthor={Taimur Khan, Jasmin Krebs, Nils Nölke, Sharad Kumar Gupta, Jonathan Renkel, Caroline Arnold (order tbd)},
  pdfkeywords={First Keyword, Second Keyword, Another Keyword.},
  colorlinks=true,
  linkcolor={black},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\hypersetup{nolinks=true}

\title{Validation challenges in large-scale tree crown segmentations
from remote sensing imagery using Deep Learning: a case study in
Germany}

\titlerunning{Validation challenges in tree crown segmentations}

\author{Taimur Khan, Jasmin Krebs, Nils Nölke, Sharad Kumar Gupta,
Jonathan Renkel, Caroline Arnold (order tbd)}

\authorrunning{T. Khan et al.}

\institute{}

\begin{document}
\maketitle

\begin{abstract}
Advances in deep learning models have opened up new avenues for computer
vision tasks, like tree crown segmentations from remote sensing images.
As more and more models become available, the challenges of applying the
models to really large spatial scales (beyond the forest/plot level)
become visible. One of these challenges is the validation of the tree
crown detection and segmentation. Imprecise detections/segmentations can
lead to inaccuracies in downstream tasks in forestry and urban planning,
creating a barrier for the application of these models to real-life use
cases. Attributes of training data, e.g tree sizes, phenological stages,
number of polygon vertices, shadows, impact the prediction accuracy of
the segmentation models, hence training data should be chosen with due
diligence. Here we identify some of the key validation challenges based
on the collective experiences of not just training deep learning models,
but also applying the models to very large spatial scales in Germany
(federal state level).

\keywords{First Keyword \and Second Keyword \and Another Keyword.}

\end{abstract}

\section{Introduction}\label{introduction}

Accurately mapping tree crowns over large areas is critical for scaling
local observations to regional and global assessments of forest health,
carbon stocks, and urban greenery. In an era marked by rapid
environmental change, comprehensive mapping efforts are indispensable.
Tree crown segmentation is a cornerstone task for these objectives, as
the size and structure of a tree crown---shaped by species-specific
branching patterns, site conditions, and competition for
light---directly influences primary production. A key metric in this
context is the crown projection area (CPA), defined as the vertical
projection of the crown onto a horizontal plane. Deriving CPA provides
essential insights both at the individual tree level---where it provides
predictions of diameter, volume, and growth rates---and at the stand
level, supporting models of competition and canopy gap dynamics.
However, inaccuracies inherent in the segmentation process, particularly
when applied at large scales, can lead to biased estimations of these
critical tree variables, underscoring the urgent need to address robust
validation challenges in large-scale tree crown segmentation. While
recent advances in deep learning offer unprecedented capabilities for
automated segmentation, their robust validation at scale presents a
critical, unaddressed hurdle.

Among the various tree crown variables, crown spread presents unique
estimation challenges, both in field measurements and, critically, when
derived from remote sensing imagery. Its accurate derivation is not only
susceptible to inaccuracies in the initial segmentation process but also
relies heavily on the methods employed to extract crown spread from
segmented polygons, a process fraught with its own complexities.
Selecting an appropriate crown spread calculation method is crucial,
especially for irregularly shaped tree crown polygons. Even with
optimized calculation methods, direct matches with ground-truth data
proved low (e.g., 32\% in our experience), although a broader acceptable
margin of error (±5 m) captured 89\% of cases. These results immediately
point to difficulties in establishing precise validation metrics.
Furthermore, model performance exhibits size-dependent biases: trees
with small crowns (\textless6 m) are often overestimated, while large
crowns (\textgreater16 m) tend to be underestimated. These discrepancies
underscore the inherent limitations of deep learning segmentation models
at the extremes of the size spectrum. Critically, the quality of
validation data profoundly influences the assessment of crown spread
accuracy. For instance, our experience with the local tree inventory
(`Baumkataster' Halle/Saale), which provides only broad crown spread
ranges (5 m intervals), severely constrains the precision of validation,
highlighting the need for more granular and reliable ground-truth
datasets.

The integration of deep learning into remote sensing has opened new ways
in how we monitor and analyse ecological systems
{[}\cite{zheng2024review}, \cite{zhao2023systematic}{]}. Tree crown
segmentation, a task central to assessing forest health, carbon stocks,
and urban greenery, has especially benefited from advances in
convolutional neural networks (CNNs) and transformer-based
architectures. Approaches leveraging high-resolution imagery have
demonstrated remarkable capabilities in delineating individual tree
crowns, as evidenced by works such as \cite{weinstein2019individual},
\cite{khant2025} and \cite{freudenberg2022individual}, illustrating the
potential for scalable and automated tree mapping. The urgency of
climate change and biodiversity loss further amplifies the importance of
large-scale tree crown segmentation. Forests serve as critical carbon
sinks and biodiversity reservoirs {[}\cite{pan2024enduring}{]}, while
urban trees provide essential ecosystem services
{[}\cite{sharma2024urban}{]}. Large-scale segmentation enables
consistent, detailed monitoring of these vital resources, informing
sustainable forest management, urban planning, and policy development.
The ambition to create national and global inventories, such as those
envisioned in \cite{tolan2024very}, hinges on the reliability of such
foundational segmentation data.

However, while these deep learning models demonstrate impressive
performance in controlled environments or with limited-scale datasets,
their deployment at expansive spatial extents, such as regional or
national scales, exposes a distinct spectrum of challenges that remain
largely unaddressed. Foremost among these is validation---a critical yet
underexplored hurdle in ensuring the reliability and generalizability of
model outputs. As recent studies have shown, the performance of tree
crown segmentation models is highly sensitive to the characteristics of
the training data, such as tree size, species diversity, seasonal
variations, and image quality {[}\cite{moussaid2021tree};
\cite{cong2022citrus}{]}. These factors, when coupled with the
complexities of real-world landscapes, underscore the need for robust
and scalable validation frameworks.

This Perspective builds on our extensive experience in applying deep
learning models to large-scale tree crown segmentation across Germany,
addressing federal state-level landscapes. By synthesizing insights from
model training, deployment, and validation, we highlight the nuanced
challenges associated with assessing segmentation accuracy over
expansive and heterogeneous terrains. These challenges are not mere
technicalities---they bear significant implications for downstream
applications in forestry, conservation, and urban planning. Through this
work, we aim to not only highlight these critical validation hurdles but
also to foster a dialogue on the methodological innovations required to
ensure that deep learning advancements in remote sensing translate into
actionable insights at the scales demanded by global environmental
challenges.

\section{Case Study: Tree Crown Segmentation in Saxony and
Saxony-Anhalt,
Germany}\label{case-study-tree-crown-segmentation-in-saxony-and-saxony-anhalt-germany}

In this case study, the DeepTrees model {[}\cite{khan2025}{]} was
applied in a one-shot prediction approach using pretrained model weights
provided by DeepTrees, applied to high-resolution multispectral digital
orthophoto imagery (DOP20) covering the German federal states of Saxony
(SN) and Saxony-Anhalt (ST). Sachsen, covering approximately 18,450 km²,
and Saxony-Anhalt, spanning around 20,452 km², represent diverse
ecological and urban landscapes, ideal for assessing large-scale
segmentation model performance.

DOP20 imagery provides detailed spatial information at 20 cm resolution
per pixel, capturing visual and near infrared spectral regions essential
for accurate tree crown delineation. Using these pretrained weights, the
DeepTrees model identified approximately 218.7 million individual tree
crowns---137.3 million in Saxony and 81.4 million in Saxony-Anhalt. The
resulting segmentation dataset has been made publicly available on
Zenodo (https://zenodo.org/record/exampledoi), facilitating further
research and validation.

@@

\emph{Figure 1: 6 panels (1 for each state). From left to right
-\textgreater{} 1) tiles with date classification 2) Land use (CORINE) +
isolines. 3) the tree segmentations.}

\section{Challenges}\label{challenges}

Validating large-scale tree crown segmentation models reveals a web of
interrelated challenges that go beyond those encountered in small,
controlled datasets. Seasonal phenology is a moving target: the same
forest can look drastically different between leaf-on summer imagery and
leaf-off winter scenes. Models trained on one season often struggle in
another, yielding inconsistent segmentation accuracy across the year.
For example, a canopy delineation that performs well on lush summer
foliage may under-segment sparse autumn crowns or miss bare branches in
winter. Such season-driven variability not only degrades model
performance but also complicates validation -- a one-shot model might
appear accurate in one season and fail in the next, raising questions
about how and when accuracy should be assessed. Incorporating
multi-season data during both training and validation is essential, as
phenological dynamics have been shown to strongly influence model
generalizability {[}\cite{moussaid2021tree}; \cite{cong2022citrus}{]}.

Spatial heterogeneity of landscapes, including terrain and illumination
differences, poses another major hurdle for both segmentation and its
validation. An algorithm that segments tree crowns flawlessly in a flat,
well-lit park may stumble in a shadow-drenched valley or on a steep
mountainside. Variations in ground elevation and slope alter the angle
of solar illumination, leading to uneven lighting and shadows that can
confuse models. In mountainous or rugged terrains, trees on north-facing
slopes might appear darker or partially occluded compared to those on
south-facing slopes with direct sun, even if they are the same species
and healthy. Such effects result in site-specific performance: models
often need fine-tuning or can suffer accuracy drops when moved to new
topographies or sensor angles. Weinstein et al.~(2020) observed this
kind of cross-site performance gap, where a tree detection model trained
in one region underperformed when applied to a different region's
imagery without adaptation, underscoring how terrain and context
influence outcomes {[}\cite{weinstein2020cross}{]}. For validation, this
means that accuracy estimates from one area may not transfer to another
-- a critical issue when assessments at national or global scales are
required.

A further fundamental challenge lies in the scarcity of accurate
ground-truth data at scale. Reliable validation hinges on high-quality
reference data (the ``ground truth''), yet collecting detailed crown
delineations over large regions is logistically difficult and expensive.
Field surveys can map individual trees with great precision (e.g.,
measuring trunks and canopy spread on the ground), but doing this over
thousands of square kilometers is infeasible. Aerial and satellite
imagery provide broader coverage for reference data, but even these
often lack the resolution or fidelity to unequivocally label each tree
crown for validation purposes. UAV (drone) campaigns can bridge the gap
by capturing very high-resolution images or LiDAR of sample areas, but
they are limited in flight range and still require extensive human
annotation to turn imagery into usable ground truth. The net result is a
mismatch of scales: our models aspire to map every tree across entire
countries, but our ground truth typically covers only small plots or
scattered samples {[}\^{}{]}. This mismatch means that validating a
``wall-to-wall'' tree map often involves extrapolating from a tiny
fraction of ground-referenced trees, introducing uncertainty. Moreover,
ground-reference datasets may not capture the full diversity of
conditions (species, canopy shapes, management regimes, etc.) present in
the larger mapping area, biasing the validation. Expanding ground-truth
collection -- through automated methods or crowdsourcing -- is thus not
just a recommendation but a necessity to overcome this validation
bottleneck (as we discuss later).

Compounding the issue of limited data is the inconsistency in reference
annotations and evaluation metrics. Even when ground-reference data
exist, their format can differ -- sometimes reference trees are marked
by a single GPS point (e.g., trunk location), sometimes by a hand-drawn
polygon outlining the crown. This creates a challenge in validation: how
do we decide if a predicted crown polygon ``matches'' a ground-truth
point, or how to handle cases where one field-mapped tree corresponds to
multiple overlapping crown segments in the image?(Figure ref)
Conversely, field crews might delineate a broad canopy as one crown
while an automated model splits it into two segments (or vice versa),
especially in dense stands where crowns merge. These ambiguities in
one-to-one correspondence make it hard to define what a ``correct''
segmentation is. Traditional pixel-wise accuracy metrics like
Intersection-over-Union (IoU) treat segmentation purely as an image
overlap problem, which may not reflect the ecological reality of
counting individual trees. IoU penalizes differences in shape or area
but doesn't account for whether the count of tree objects is correct. In
an extreme case, a model could slightly over-segment every tree
(splitting each true crown into two smaller polygons) and still achieve
a reasonable IoU, despite doubling the perceived tree count -- a
significant error for applications. On the other hand, object-centric
metrics such as panoptic segmentation quality attempt to consider both
detection and delineation of objects
{[}\cite{kirillov2019panopticsegmentation}{]}. Panoptic metrics combine
aspects of object detection (was each tree detected?) with segmentation
quality (was each crown correctly outlined?), which can be more
appropriate for tree mapping. However, even these require well-defined
ground-truth objects to compare against. When the ground truth itself is
inconsistent (e.g., how to count a clumped cluster of stems with
overlapping crowns), validation metrics struggle to fully capture model
performance. The choice of evaluation metric thus becomes non-trivial:
depending on whether one prioritizes exact crown shape, tree count, or
canopy cover, the ``best'' metric may differ. Establishing consensus on
evaluation protocols is part of the challenge -- without it, different
studies may report accuracy in incompatible ways.

There is also the issue of scale and resolution in validation reporting.
A model's accuracy can appear to vary depending on the spatial scale at
which it's evaluated. For instance, a segmentation model might achieve
high overall accuracy when averaged over an entire region, yet if one
zooms into a small test area (say a single city park or forest stand),
the error rate might be much higher or lower. This can happen if errors
are not evenly distributed: the model could perform very well in one
type of landscape (e.g., neat urban street trees) and poorly in another
(dense natural forest), and a coarse regional average could mask these
extremes. Consequently, a user working on a local conservation project
might experience worse performance than the ``headline'' accuracy
suggests, because that headline number was diluted by many easier cases
elsewhere. Ensuring that validation is robust across scales is tricky --
one must balance broad coverage with local detail. It calls for
multi-scale validation approaches, where accuracy is reported at
multiple grain sizes or stratified by landscape type. Highlighting this,
one could imagine a figure plotting model accuracy as a function of
spatial extent or across different habitat types, illustrating how
performance can drop off in specific challenging subsets despite looking
good overall. This emphasizes that heterogeneous performance is itself a
challenge to acknowledge in validation: we need methods to detect where
and why a model fails, not just an aggregate score.

To summarize, validating large-scale tree segmentation models is far
from a straightforward task. Seasonal changes, diverse terrain and
illumination, limited and inconsistent ground truth, ambiguous
evaluation criteria, and scale-dependent performance all intertwine to
create a demanding setting. These challenges are not just academic --
they directly impact how much trust we can place in AI-generated tree
maps for real-world decisions in forestry, ecology, and urban planning.
Recognizing these pain points is the first step; the next is devising
strategies to overcome them. We now turn to several recommendations
aimed at improving both models and validation practices, with an eye
toward bridging the gap between controlled experiments and the
complexity of continental-scale deployments.

\emph{Table 1: Potential validation challenges and corresponding
mitigation strategies. This table could list key challenges in
validating large-scale tree segmentations alongside proposed approaches
to address them. For example: (Challenge) Phenological variation between
seasons -- (Mitigation) use multi-temporal training data and
season-specific validation sets; (Challenge) Heterogeneous terrain and
shadowing -- (Mitigation) integrate digital elevation models and
terrain-aware modeling; (Challenge) Scarcity of ground-truth annotations
-- (Mitigation) leverage semi-automated labeling, crowdsourcing, and
active learning to expand validation data; (Challenge) Ambiguity in
crown delineation -- (Mitigation) adopt object-based accuracy metrics
and consensus protocols for what counts as a single tree; (Challenge)
Scale-dependent performance -- (Mitigation) evaluate models at multiple
spatial scales and stratify results by landscape type. This overview
would visually reinforce the narrative that each challenge has an
identifiable path forward, linking the ``Challenges'' and
``Recommendations'' sections.}

\emph{!{[}{]}{[}image1{]}}\\
\emph{Figure 2: Incomplete}

\section{Recommendations}\label{recommendations}

>>>>

---
title: "Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany"
title-running: Validation challenges in tree crown segmentations
format:
  quarto-ecml-pdf:
    keep-tex: true
  quarto-ecml-html: default
author: Taimur Khan, Jasmin Krebs, Nils Nölke, Sharad Kumar Gupta, Jonathan Renkel, Caroline Arnold (order tbd)
author-running: T. Khan et al.
abstract: |
  Advances in deep learning models have opened up new avenues for computer vision tasks, like tree crown segmentations from remote sensing images. As more and more models become available, the challenges of applying the models to really large spatial scales (beyond the forest/plot level) become visible. One of these challenges is the validation of the tree crown detection and segmentation. Imprecise detections/segmentations can lead to inaccuracies in downstream tasks in forestry and urban planning, creating a barrier for the application of these models to real-life use cases. Attributes of training data, e.g tree sizes, phenological stages, number of polygon vertices, shadows, impact the prediction accuracy of the segmentation models, hence training data should be chosen with due diligence. Here we identify some of the key validation challenges based on the collective experiences of not just training deep learning models, but also applying the models to very large spatial scales in Germany (federal state level).
keywords: [First Keyword, Second Keyword, Another Keyword.]
---

## Introduction

Accurately mapping tree crowns over large areas is critical for scaling local observations to regional and global assessments of forest health, carbon stocks, and urban greenery. In an era marked by rapid environmental change, comprehensive mapping efforts are indispensable. Tree crown segmentation is a cornerstone task for these objectives, as the size and structure of a tree crown—shaped by species-specific branching patterns, site conditions, and competition for light—directly influences primary production. A key metric in this context is the crown projection area (CPA), defined as the vertical projection of the crown onto a horizontal plane. Deriving CPA provides essential insights both at the individual tree level—where it provides predictions of diameter, volume, and growth rates—and at the stand level, supporting models of competition and canopy gap dynamics. However, inaccuracies inherent in the segmentation process, particularly when applied at large scales, can lead to biased estimations of these critical tree variables, underscoring the urgent need to address robust validation challenges in large-scale tree crown segmentation. While recent advances in deep learning offer unprecedented capabilities for automated segmentation, their robust validation at scale presents a critical, unaddressed hurdle.

Among the various tree crown variables, crown spread presents unique estimation challenges, both in field measurements and, critically, when derived from remote sensing imagery. Its accurate derivation is not only susceptible to inaccuracies in the initial segmentation process but also relies heavily on the methods employed to extract crown spread from segmented polygons, a process fraught with its own complexities. Selecting an appropriate crown spread calculation method is crucial, especially for irregularly shaped tree crown polygons. Even with optimized calculation methods, direct matches with ground-truth data proved low (e.g., 32% in our experience), although a broader acceptable margin of error (±5 m) captured 89% of cases. These results immediately point to difficulties in establishing precise validation metrics. Furthermore, model performance exhibits size-dependent biases: trees with small crowns (\<6 m) are often overestimated, while large crowns (\>16 m) tend to be underestimated. These discrepancies underscore the inherent limitations of deep learning segmentation models at the extremes of the size spectrum. Critically, the quality of validation data profoundly influences the assessment of crown spread accuracy. For instance, our experience with the local tree inventory ('Baumkataster' Halle/Saale), which provides only broad crown spread ranges (5 m intervals), severely constrains the precision of validation, highlighting the need for more granular and reliable ground-truth datasets.

The integration of deep learning into remote sensing has opened new ways in how we monitor and analyse ecological systems \[@zheng2024review, @zhao2023systematic\]. Tree crown segmentation, a task central to assessing forest health, carbon stocks, and urban greenery, has especially benefited from advances in convolutional neural networks (CNNs) and transformer-based architectures. Approaches leveraging high-resolution imagery have demonstrated remarkable capabilities in delineating individual tree crowns, as evidenced by works such as @weinstein2019individual, @khant2025 and @freudenberg2022individual, illustrating the potential for scalable and automated tree mapping. The urgency of climate change and biodiversity loss further amplifies the importance of large-scale tree crown segmentation. Forests serve as critical carbon sinks and biodiversity reservoirs \[@pan2024enduring\], while urban trees provide essential ecosystem services \[@sharma2024urban\]. Large-scale segmentation enables consistent, detailed monitoring of these vital resources, informing sustainable forest management, urban planning, and policy development. The ambition to create national and global inventories, such as those envisioned in @tolan2024very, hinges on the reliability of such foundational segmentation data.

However, while these deep learning models demonstrate impressive performance in controlled environments or with limited-scale datasets, their deployment at expansive spatial extents, such as regional or national scales, exposes a distinct spectrum of challenges that remain largely unaddressed. Foremost among these is validation—a critical yet underexplored hurdle in ensuring the reliability and generalizability of model outputs. As recent studies have shown, the performance of tree crown segmentation models is highly sensitive to the characteristics of the training data, such as tree size, species diversity, seasonal variations, and image quality \[@moussaid2021tree; @cong2022citrus\]. These factors, when coupled with the complexities of real-world landscapes, underscore the need for robust and scalable validation frameworks.

This Perspective builds on our extensive experience in applying deep learning models to large-scale tree crown segmentation across Germany, addressing federal state-level landscapes. By synthesizing insights from model training, deployment, and validation, we highlight the nuanced challenges associated with assessing segmentation accuracy over expansive and heterogeneous terrains. These challenges are not mere technicalities—they bear significant implications for downstream applications in forestry, conservation, and urban planning. Through this work, we aim to not only highlight these critical validation hurdles but also to foster a dialogue on the methodological innovations required to ensure that deep learning advancements in remote sensing translate into actionable insights at the scales demanded by global environmental challenges.

## Case Study: Tree Crown Segmentation in Saxony and Saxony-Anhalt, Germany

In this case study, the DeepTrees model \[@khan2025\] was applied in a one-shot prediction approach using pretrained model weights provided by DeepTrees, applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). Sachsen, covering approximately 18,450 km², and Saxony-Anhalt, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance.

DOP20 imagery provides detailed spatial information at 20 cm resolution per pixel, capturing visual and near infrared spectral regions essential for accurate tree crown delineation. Using these pretrained weights, the DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in Saxony and 81.4 million in Saxony-Anhalt. The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi), facilitating further research and validation.

@@

*Figure 1: 6 panels (1 for each state). From left to right \-\> 1\) tiles with date classification 2\) Land use (CORINE) \+ isolines. 3\) the tree segmentations.*

## Challenges

Validating large-scale tree crown segmentation models reveals a web of interrelated challenges that go beyond those encountered in small, controlled datasets. Seasonal phenology is a moving target: the same forest can look drastically different between leaf-on summer imagery and leaf-off winter scenes. Models trained on one season often struggle in another, yielding inconsistent segmentation accuracy across the year. For example, a canopy delineation that performs well on lush summer foliage may under-segment sparse autumn crowns or miss bare branches in winter. Such season-driven variability not only degrades model performance but also complicates validation – a one-shot model might appear accurate in one season and fail in the next, raising questions about how and when accuracy should be assessed. Incorporating multi-season data during both training and validation is essential, as phenological dynamics have been shown to strongly influence model generalizability \[@moussaid2021tree; @cong2022citrus\].

Spatial heterogeneity of landscapes, including terrain and illumination differences, poses another major hurdle for both segmentation and its validation. An algorithm that segments tree crowns flawlessly in a flat, well-lit park may stumble in a shadow-drenched valley or on a steep mountainside. Variations in ground elevation and slope alter the angle of solar illumination, leading to uneven lighting and shadows that can confuse models. In mountainous or rugged terrains, trees on north-facing slopes might appear darker or partially occluded compared to those on south-facing slopes with direct sun, even if they are the same species and healthy. Such effects result in site-specific performance: models often need fine-tuning or can suffer accuracy drops when moved to new topographies or sensor angles. Weinstein et al. (2020) observed this kind of cross-site performance gap, where a tree detection model trained in one region underperformed when applied to a different region’s imagery without adaptation, underscoring how terrain and context influence outcomes \[@weinstein2020cross\]. For validation, this means that accuracy estimates from one area may not transfer to another – a critical issue when assessments at national or global scales are required.

A further fundamental challenge lies in the scarcity of accurate ground-truth data at scale. Reliable validation hinges on high-quality reference data (the “ground truth”), yet collecting detailed crown delineations over large regions is logistically difficult and expensive. Field surveys can map individual trees with great precision (e.g., measuring trunks and canopy spread on the ground), but doing this over thousands of square kilometers is infeasible. Aerial and satellite imagery provide broader coverage for reference data, but even these often lack the resolution or fidelity to unequivocally label each tree crown for validation purposes. UAV (drone) campaigns can bridge the gap by capturing very high-resolution images or LiDAR of sample areas, but they are limited in flight range and still require extensive human annotation to turn imagery into usable ground truth. The net result is a mismatch of scales: our models aspire to map every tree across entire countries, but our ground truth typically covers only small plots or scattered samples \[^\]. This mismatch means that validating a “wall-to-wall” tree map often involves extrapolating from a tiny fraction of ground-referenced trees, introducing uncertainty. Moreover, ground-reference datasets may not capture the full diversity of conditions (species, canopy shapes, management regimes, etc.) present in the larger mapping area, biasing the validation. Expanding ground-truth collection – through automated methods or crowdsourcing – is thus not just a recommendation but a necessity to overcome this validation bottleneck (as we discuss later).

Compounding the issue of limited data is the inconsistency in reference annotations and evaluation metrics. Even when ground-reference data exist, their format can differ – sometimes reference trees are marked by a single GPS point (e.g., trunk location), sometimes by a hand-drawn polygon outlining the crown. This creates a challenge in validation: how do we decide if a predicted crown polygon “matches” a ground-truth point, or how to handle cases where one field-mapped tree corresponds to multiple overlapping crown segments in the image?(Figure ref)  Conversely, field crews might delineate a broad canopy as one crown while an automated model splits it into two segments (or vice versa), especially in dense stands where crowns merge. These ambiguities in one-to-one correspondence make it hard to define what a “correct” segmentation is. Traditional pixel-wise accuracy metrics like Intersection-over-Union (IoU) treat segmentation purely as an image overlap problem, which may not reflect the ecological reality of counting individual trees. IoU penalizes differences in shape or area but doesn’t account for whether the count of tree objects is correct. In an extreme case, a model could slightly over-segment every tree (splitting each true crown into two smaller polygons) and still achieve a reasonable IoU, despite doubling the perceived tree count – a significant error for applications. On the other hand, object-centric metrics such as panoptic segmentation quality attempt to consider both detection and delineation of objects \[@kirillov2019panopticsegmentation\]. Panoptic metrics combine aspects of object detection (was each tree detected?) with segmentation quality (was each crown correctly outlined?), which can be more appropriate for tree mapping. However, even these require well-defined ground-truth objects to compare against. When the ground truth itself is inconsistent (e.g., how to count a clumped cluster of stems with overlapping crowns), validation metrics struggle to fully capture model performance. The choice of evaluation metric thus becomes non-trivial: depending on whether one prioritizes exact crown shape, tree count, or canopy cover, the “best” metric may differ. Establishing consensus on evaluation protocols is part of the challenge – without it, different studies may report accuracy in incompatible ways.

There is also the issue of scale and resolution in validation reporting. A model’s accuracy can appear to vary depending on the spatial scale at which it’s evaluated. For instance, a segmentation model might achieve high overall accuracy when averaged over an entire region, yet if one zooms into a small test area (say a single city park or forest stand), the error rate might be much higher or lower. This can happen if errors are not evenly distributed: the model could perform very well in one type of landscape (e.g., neat urban street trees) and poorly in another (dense natural forest), and a coarse regional average could mask these extremes. Consequently, a user working on a local conservation project might experience worse performance than the “headline” accuracy suggests, because that headline number was diluted by many easier cases elsewhere. Ensuring that validation is robust across scales is tricky – one must balance broad coverage with local detail. It calls for multi-scale validation approaches, where accuracy is reported at multiple grain sizes or stratified by landscape type. Highlighting this, one could imagine a figure plotting model accuracy as a function of spatial extent or across different habitat types, illustrating how performance can drop off in specific challenging subsets despite looking good overall. This emphasizes that heterogeneous performance is itself a challenge to acknowledge in validation: we need methods to detect where and why a model fails, not just an aggregate score.

To summarize, validating large-scale tree segmentation models is far from a straightforward task. Seasonal changes, diverse terrain and illumination, limited and inconsistent ground truth, ambiguous evaluation criteria, and scale-dependent performance all intertwine to create a demanding setting. These challenges are not just academic – they directly impact how much trust we can place in AI-generated tree maps for real-world decisions in forestry, ecology, and urban planning. Recognizing these pain points is the first step; the next is devising strategies to overcome them. We now turn to several recommendations aimed at improving both models and validation practices, with an eye toward bridging the gap between controlled experiments and the complexity of continental-scale deployments.

*Table 1: Potential validation challenges and corresponding mitigation strategies. This table could list key challenges in validating large-scale tree segmentations alongside proposed approaches to address them. For example: (Challenge) Phenological variation between seasons – (Mitigation) use multi-temporal training data and season-specific validation sets; (Challenge) Heterogeneous terrain and shadowing – (Mitigation) integrate digital elevation models and terrain-aware modeling; (Challenge) Scarcity of ground-truth annotations – (Mitigation) leverage semi-automated labeling, crowdsourcing, and active learning to expand validation data; (Challenge) Ambiguity in crown delineation – (Mitigation) adopt object-based accuracy metrics and consensus protocols for what counts as a single tree; (Challenge) Scale-dependent performance – (Mitigation) evaluate models at multiple spatial scales and stratify results by landscape type. This overview would visually reinforce the narrative that each challenge has an identifiable path forward, linking the “Challenges” and “Recommendations” sections.*

*![][image1]*  
*Figure 2: Incomplete*

## Recommendations
